Cosmological Evidences of Dark Matter through the CMB

### 2019-02-21 paper/main.tex
sachs-wolfe effect

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[a4paper, total={6in, 8in}]{geometry}
\title{Cosmological Evidences of Dark Matter through the CMB}
\author{Lorenzo Speri}
\date{}
%\usepackage{hyperref}
\usepackage{natbib}
\usepackage{graphicx}
% ---- Maths -------
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{physics}
% ---- Links -------
%\usepackage{hyperref}
% citation color
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
% for twnsore
\usepackage{tensind}
\tensordelimiter{:}

@@

\section{Introduction}
Many astronomical and cosmological observations suggest that, by using the General Theory of Relativity, there is an invisible form of matter that exerts a gravitational pull on baryonic matter. 
In order to explain such observations, scientists have proposed the existence of Dark Matter, a new form of matter that does not interact electromagnetically and, therefore, it appears invisible to us. 
The fundamental nature of Dark Matter is not well understood yet, and whether Dark Matter is truly a new form of matter or the consequence of new laws of gravity is still matter of debate.\\
One of the most compelling evidence for the existence of Dark Matter is given by the analysis of the Cosmic Microwave Background (CMB).\\
The Cosmic Microwave Background is the electromagnetic radiation coming from the hot plasma of the early stages of the Universe.
As the Universe was expanding, it cooled down to a temperature low enough to form stable hydrogen $e^- + p^+  \rightarrow H + \gamma$ (Recombination). 
As the number density of the free electrons dropped with the expansion, the Thomson scattering $e^- + \gamma  \rightarrow e^- + \gamma$ became inefficient and the photons have since streamed freely through the universe in all directions and they are today observed as the Cosmic Microwave Background \citep{LecturesPdf}.\\
The Cosmic Microwave Background appears to us as the best measured Planck spectrum that we know with fluctuations of order $10 ^{-5}$ around the average temperature $T_{CMB} =2.725$ K.
The fluctuations are extremely important because they contain the crucial information about the primordial plasma and its content. 
In fact, during the early stages of the Universe, the photons were interacting with baryonic matter, which in turn was interacting gravitationally with dark matter.
Thus, the observations of the CMB fluctuations are linked to the energy-matter content of our Universe and dark matter.\\
The aim of this paper is to give a short and concise overview of the physics of the CMB and the cosmological evidence of Dark Matter through the analysis of the CMB fluctuations within the $\Lambda$CDM model assumptions.
We are interested in providing few but fundamental tools to understand the importance of the CMB for Dark Matter.\\
We first give an historical overview of the discovery and measurements of the CMB, then we discuss the main assumptions of the $\Lambda$CDM model and the link between the content of our universe and the cosmological model. 
Afterwards, the phenomenology of the hot plasma is explained with a simplified treatment in order to understand the power spectrum of the CMB.
We conclude with the discussion of the CMB temperature power spectrum and the role of dark matter.

%The recent measurements of the Cosmic Microwave Background (CMB) radiation allowed us to infer the presence of Dark Matter in the Universe. In this summary we will explain qualitatively how the presence of Dark Matter influences the CMB anisotropies.

\section{From the Discovery of the CMB to the Planck mission}

\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{Horn_Antenna.jpeg}
\caption{The 15 meter Holmdel horn antenna at Bell Telephone Laboratories in Holmdel, New Jersey.
(By NASA - Great Images in NASA Description, Public Domain, \url{https://commons.wikimedia.org/w/index.php?curid=6463768})}
\end{center}

\end{figure}

In 1965 Arno Penzias and Robert Wilson published the paper: \emph{A measurement of excess antenna temperature at
4080 Mc/s}, where they reported the measurements of an isotropic, unpolarized, and free from seasonal variations excess noise temperature
\citep{penziasMeasurementExcessAntenna1965}. 
When Penzias and Wilson first measured serendipitously this strong signal, they did everything they could think of to reduce “noise” in their system. 
They even shooed away a pair of pigeons that had roosted in the antenna and cleaned up they later called “the usual white dielectric” generated by pigeons \citep{RydenIntroCosmoPdf}.
However, the signal remained. 
Robert Dicke and his research group at Princeton University explained such signal as the relic of an early, hot, dense, and opaque state of the universe\footnote{the existence of the cosmic background radiation had actually been predicted by the physicist George Gamow and his colleagues in 1948} \citep{bucherPhysicsCosmicMicrowave2015}
. \\
This discovery of the CMB created a big excitement and a new era of cosmology began.
Measuring the CMB spectrum and its deviation from the black body spectrum was the new challange.
Many experimental efforts have been carried out to measure accurately the fluctuations of Cosmic Microwave Backgrund from its average temperature $T = 2.7$ K, we will discuss only few of them.
\\
First of all, the photons of the CMB can be absorbed by the Earth's atmosphere, because the energy per CMB photon (approximately $\sim 6 \times 10 ^{-4}$ eV) is comparable to the energy of vibration or rotation for a small molecule (of water for instance). In fact, microwaves with wavelengths shorter than $\lambda \sim 3$ cm are strongly absorbed by water molecules \citep{RydenIntroCosmoPdf}.
This problem could be overcomed by observing the CMB in different bands and from locations where the humidity is low.
However, the best way to meseaure the CMB spectrum is to use satellites.\\
The first sattellite that was launched to observe the CMB was COBE. 
It did provide a convincing first detection of the CMB anisotropy, and it played a crucial role in determining the viability of the different cosmological models at that time.
\\
After that, WMAP and PLanck space missions  increased the accuracy of the measurements Figure (\ref{cobe_wmap_planck}).
Nevertheless the satellites improved their accuracy, many other sources of noise were encountered such as the light coming from the center of our galaxy, other stars and other objects in the universe, in addition to the relative motion of the satellite with respect to the CMB (Figure (\ref{cobe_map})).\\
Finally, observations and statistical analysis showed that the CMB spectrum is close to a black body spectrum up to fluctuations of order of $10^{-5}$ K (Figure (\ref{cobe_blackbody})).\\
% maybe last sentence in the introduction with the image

\begin{figure}
\centering
    \textbf{Cosmic Microwave Background  Spectrum}\par\medskip
\centering
   {\includegraphics[height=8cm]{blackbody}}
\caption{}
\label{cobe_blackbody}
\end{figure}

\begin{figure}
\centering
    \textbf{The Cosmic Microwave Background anisotropies measured by COBE, WMAP and Planck}\par\medskip
\centering
   {\includegraphics[width=.75\textwidth]{cmb1.jpg}}

\caption{The COBE, WMAP and Planck missions measured the CMB anisotropies with increasing accuracy. COBE, the first CMB satellite, measured fluctuations to scales of 7$^\circ$. WMAP was able to measure resolutions down to 0.3$^\circ$ in five different frequency bands, with Planck measuring all the way down to just 5 arcminutes (0.07$^\circ$) in nine different frequency bands in total.  (NASA/COBE/DMR; NASA/WMAP SCIENCE TEAM; ESA AND THE PLANCK COLLABORATION)}
\label{cobe_wmap_planck}
\end{figure}

\begin{figure}
\centering
    \textbf{Orbits of different  configurations of BBH}\par\medskip
\centering
\includegraphics[scale =0.3]{mono_di_cobe}
\caption{The microwave sky as seen by the COBE DMR (differential microwave radiometer) instrument. The top panel shows the microwave sky as seen on a linear tem- perature scale including zero. No anisotropies are visible in this image, because the CMB monopole at 2.725K dominates. In the middle panel, the monopole component has been subtracted. Apart from some slight contamination from the galaxy near the equator (corre- sponding to the plane of our Galaxy), one sees only a nearly perfect dipole pattern, owing to our peculiar motion with respect to the rest frame defined by the CMB. In the bottom panel, both the monopole and dipole components have been removed. Except for the galactic contamination around the equator, one sees the cosmic microwave background anisotropy along with some noise. (Credit: NASA/COBE Science Team)}
\label{cobe_map}
\end{figure}

@@

In this section we explain where the components of our Universe (Dark Matter, baryonic matter, Dark Energy, curvature)  arise from and how they are related with each others. 

@@

@@

Given these assumptions, the most general metric $g_{\mu \nu}$ satisfying homogeneity and isotropy is the Friedmann–Lemaître–Robertson–Walker (FLRW) metric written here in terms of the invariant geodesic distance $\dd s ^2 = g_{\mu \nu} \dd x^{\mu} \dd x^{\nu}$:

@@

where $k$ is the spatial curvature and $a(t)$ is the scale factor.
Notice that the FLRW metric has $g_{0 i}=0$ due to the isotropy.

@@

in order to evolve the FLRW metric.
The Einstein's field equations relate how the geometry of spacetime (left hand side) changes with the presence of masses and energy (right hand side).

@@

Solving (\ref{einstein_eq}) for $\mu= 0=\nu$ using the above definition (\ref{energy-momentum-tensor}), gives us the Friedmann equation:

@@

H^2 = \qty(\frac{\dot{a}}{a})^2 = \dfrac{8 \pi G}{3} \rho -\dfrac{k \, c^2}{a^2} + \frac{\Lambda \, c^2}{3}

@@

where $\dot{a}$ is the time derivative of the scale factor.\\
And the conservation of energy momentum tensor $\nabla_\mu T^{\mu 0} =0 $ becomes:

@@

@@

Solving this for matter $\rho_m << p_m$ (or \emph{dust}) and for radiation $\rho_r = p_r/3$ gives

@@

where $\rho_{m 0}$ and $\rho_{r 0}$ are the matter density and radiation density evaluated today $t = t_0$, $a(t_0) = a_0 =1$.\\
We now rewrite equation (\ref{firedmann_eq}) as

@@

where we defined $H_0 = H(t_0)$.\\

@@

\Omega_\Lambda = \dfrac{\Lambda c^2}{3} \qquad \Omega_k = -  \dfrac{k c^2}{H_0 ^2} \qquad
\Omega_m = \dfrac{8 \pi G}{3 H_0 ^2} \rho_{m 0} \qquad \Omega_r =\dfrac{8 \pi G}{3 H_0 ^2} \rho_{r 0}

@@

evaluating the Friedmann equation (\ref{firedmann_eq}) today we obtain the cosmic sum rule

@@

\Omega_{\Lambda} + \Omega_{m} + \Omega_k + \Omega_{r} = 1

@@

\section{Phenomenology of the Fluctuations}
The Cosmic Microwave Background is the image of the fluctuations of a dense and hot fluid made of matter and radiation.
Before the release of the CMB at $z > 1100$ the electromagnetic radiation is firmly coupled (both energetically and kinetically) to the baryons and it is possible to describe the behavior of such a system by using solving the equations of relativistic hydrodynamics and the Einstein's field equations (\ref{einstein_eq}) with small density $\delta \rho$, pressure $\delta p$ and gravitational potential perturbations $\delta \Phi$.
The full general-relativistic description of the photon-matter fluid requires a complicated treatment[BAUMANN].
However, it is possible to qualitatively discuss the phenomena occurring in the CMB fluid and the effects on the temperature fluctuations $\Theta = \delta T / T_0$.
\subsection{Sachs-Wolfe effect}
As a consequence of the Einstein's equivalence principle, photons that escape from a gravitational field are redshifted.
If we consider the fluctuations of the CMB on large scales, the photons that tries to escape from the gravitational well created by matter are redshifted.

@@

 by starting from the continuity and Euler equation for a photon fluid with  number density $n$, energy density $u$ and pressure $p$:
\begin{equation}
\label{hydro}
\dot{n} + n_0 \vec{\nabla} \cdot \vec{v} = 0 \qquad
\dot{\vec{v}} = - c^2 \dfrac{\vec{\nabla} \delta p}{p_0 + u_0} - \vec{\nabla} \delta \Phi
\end{equation}
where $v$ is the streaming velocity of the perturbations and the subscript $_0$ indicates the background average values of th correspondent quantities.\\
For a photon gas we know that
\begin{equation}
n \propto T ^3 \qquad u \propto T^4 \qquad p  = \dfrac{u}{3}
\end{equation}
So, the temperature fluctuations are related to the number density, energy density and pressure as follow:
\begin{equation}
\dfrac{\delta n}{n_0} = 3 \Theta \qquad
\dfrac{\delta u}{u_0} = 4 \Theta = \dfrac{\delta p}{p_0}
\end{equation}
We can rewrite eq.(\ref{hydro}) in terms of the temperature fluctuations:
\begin{equation}
\dot{\Theta} + \nabla \cdot \vec{v} = 0 \qquad \dot{\vec{v}} = - c^2 \vec{\nabla} \Theta - \vec{\nabla} \delta \Phi
\end{equation}
where we used $p_0 + u_0 = 4 p_0$

@@

\section{Comments on the results and plots of the CMB spectrum}
We have seen in section 3 how general relativity and the density parameters  characterize the content of our Universe,  in section 4 the dynamics of the CMB photons is affected by matter.
We want to link the previous theoretical results to the observations of the temperature fluctuations of the CMB sky.\\
The spherical harmonics $Y^{m} _{l} (\theta, \phi)$ compose a set of orthonormal functions on the sky.\\
So if $T(\theta,\phi)$ is the temperature at position $(\theta,\phi)$, it is possible to decompose the temperature into a series:
\begin{equation}
T(\theta, \phi) = \sum _{l, m} a_{lm} Y^{m} _l (\theta, \phi)
\end{equation}
Due to the orthonormality of the spherical harmonics
\begin{equation}
a_{lm} = \int _0 ^{2 \pi} \dd \phi \int _0  ^{\pi} \dd \theta \,
\sin \theta \, T(\theta, \phi) \, Y^{m} _l (\theta, \phi)
 \end{equation}
 We define the power spectrum of the temperature map as 
 \begin{equation}
 C_l = \dfrac{1}{2l+1} \sum _{m=-l} ^{l} |a_{lm}|^2
 \end{equation}
The measured temperature power spectrum is shown in Figure(\ref{temp_pow_spect})
%
%
%
\begin{figure}
\includegraphics[width=0.9\textwidth]{planck2018}
\caption{Planck 2018 temperature power spectrum. $\mathcal{D}^{TT}_l = l(l+1)C_l$ (arXiv:1807.06209v1)}
\label{temp_pow_spect}
\end{figure}
%
%
%
The Sachs-Wolfe effect is dominant on large scales, i.e. small l.\\
The first peak is at $l \approx 200$ and the  acoustic oscillations take place.
The damping effect is evident for $l>10^{3}$.\\

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage{geometry}
 \geometry{
 a4paper,
 bottom=40mm
 %total={170mm,257mm},
 %left=20mm,
 %top=20mm,
 }
 \title{Cosmological Evidences of Dark Matter through the CMB}
\author{Lorenzo Speri}
\date{}
%\usepackage{hyperref}
\usepackage{natbib}
\usepackage{graphicx}
% ---- Maths -------
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{physics}
% ---- Links -------
%\usepackage{hyperref}
% citation color
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
% for twnsore
\usepackage{tensind}
\tensordelimiter{:}

@@

\vspace{3cm}
\section{Introduction}
Many astronomical and cosmological observations suggest that, by using the General Theory of Relativity, there is an invisible form of matter that exerts a gravitational pull on baryonic matter. 
In order to explain such observations, scientists have proposed the existence of Dark Matter, a new form of matter that does not interact electromagnetically and, therefore, it appears invisible to us. 
The fundamental nature of Dark Matter is not well understood yet, and whether Dark Matter is truly a new form of matter or the consequence of new laws of gravity is still matter of debate.\\
One of the most compelling evidence for the existence of Dark Matter is given by the analysis of the Cosmic Microwave Background (CMB).\\
The Cosmic Microwave Background is the electromagnetic radiation coming from the hot plasma of the early stages of the Universe.
As the Universe was expanding, it cooled down to a temperature low enough to form stable hydrogen $e^- + p^+  \rightarrow H + \gamma$ (Recombination). 
As the number density of the free electrons dropped with the expansion, the Thomson scattering $e^- + \gamma  \rightarrow e^- + \gamma$ became inefficient and the photons have since streamed freely through the universe in all directions and they are today observed as the Cosmic Microwave Background \citep{LecturesPdf}.\\
The Cosmic Microwave Background appears to us as the best measured Planck spectrum that we know with fluctuations of order $10 ^{-5}$ around the average temperature $T_{CMB} =2.725$ K.
The fluctuations are extremely important because they contain the crucial information about the primordial plasma and its content. 
In fact, during the early stages of the Universe, the photons were interacting with baryonic matter, which in turn was interacting gravitationally with dark matter.
Thus, the observations of the CMB fluctuations are linked to the energy-matter content of our Universe and dark matter.\\
The aim of this paper is to give a short and concise overview of the physics of the CMB and the cosmological evidence of Dark Matter through the analysis of the CMB fluctuations within the $\Lambda$CDM model assumptions.
We are interested in providing few but fundamental tools to understand the importance of the CMB for Dark Matter.\\
We first give an historical overview of the discovery and measurements of the CMB, then we discuss the main assumptions of the $\Lambda$CDM model and the link between the content of our universe and the cosmological model. 
Afterwards, the phenomenology of the hot plasma is explained with a simplified treatment in order to understand the power spectrum of the CMB.
We conclude with the discussion of the CMB temperature power spectrum and the role of dark matter.

%The recent measurements of the Cosmic Microwave Background (CMB) radiation allowed us to infer the presence of Dark Matter in the Universe. In this summary we will explain qualitatively how the presence of Dark Matter influences the CMB anisotropies.
\pagebreak
\section{From the Discovery of the CMB to the Planck mission}
%
%
%
\begin{figure}
\begin{center}
\includegraphics[width=0.7\textwidth]{Horn_Antenna.jpeg}
\caption{The 15 meter Holmdel horn antenna at Bell Telephone Laboratories in Holmdel, New Jersey.
(By NASA - Great Images in NASA Description, Public Domain, \url{https://commons.wikimedia.org/w/index.php?curid=6463768})}
\label{antenna}
\end{center}
\end{figure}
%
%
%
In the 1960s Penzias and Wilson measured an isotropic and constant with time strong signal coming from the sky by using a horn-reflector radio antenna (Figure(\ref{antenna})). 
They tried everything they could do to reduce the “noise” in their system, but the signal remained.\\
In 1965 Arno Penzias and Robert Wilson published the paper: \emph{A measurement of excess antenna temperature at
4080 Mc/s}, where they reported the measurements of an isotropic, unpolarized, and free from seasonal variations excess noise temperature
\citep{penziasMeasurementExcessAntenna1965}.
The explanation of that signal was given by Robert Dicke and his research group at Princeton University: such signal as the relic of an early, hot, dense, and opaque state of the universe\footnote{The existence of the cosmic background radiation had actually been predicted by the physicist George Gamow and his colleagues in 1948} \citep{bucherPhysicsCosmicMicrowave2015}
. \\
The discovery of the CMB created a big excitement and a new era of cosmology began.
Measuring the CMB spectrum and its deviation from the black body spectrum was the new challange.
\\
First of all, the photons of the CMB can be absorbed by the Earth's atmosphere, because the energy per CMB photon (approximately $\sim 6 \times 10 ^{-4}$ eV) is comparable to the energy of vibration or rotation for a small molecule (of water for instance). 
In fact, microwaves with wavelengths shorter than $\lambda \sim 3$ cm are strongly absorbed by water molecules \citep{RydenIntroCosmoPdf}.
This problem could be overcomed by observing the CMB in different bands and from locations where the atmospheric humidity is low, at high altitudes and at low temperatures.
However, the best way to meseaure the CMB spectrum is to use satellites.\\
The first sattellite that was launched to observe the CMB was COBE. 
It did provide a convincing first detection of the CMB anisotropy, and it played a crucial role in determining the viability of the different cosmological models at that time.
\\
After that, WMAP and PLanck space missions  increased the accuracy of the measurements Figure (\ref{cobe_wmap_planck}).\\
The COBE, WMAP and Planck missions measured the CMB anisotropies with increasing accuracy. COBE, the first CMB satellite, measured fluctuations to scales of 7$^\circ$. WMAP was able to measure resolutions down to 0.3$^\circ$ in five different frequency bands, with Planck measuring all the way down to just 5 arcminutes (0.07$^\circ$) in nine different frequency bands in total.\\
Nevertheless the satellites improved their accuracy, many other sources of noise were encountered such as the light coming from the center of our galaxy, other stars and other objects in the universe, in addition to the relative motion of the satellite with respect to the CMB (Figure (\ref{cobe_map})).\\ 
Finally, observations and statistical analysis showed that the CMB spectrum is close to a black body spectrum up to fluctuations of order of $10^{-5}$ K (Figure (\ref{cobe_blackbody})).\\
% maybe last sentence in the introduction with the image

\begin{figure}[h]
\centering
    \textbf{Cosmic Microwave Background  Spectrum}\par\medskip
\centering
   {\includegraphics[height=7cm]{blackbody}}
\caption{}
\label{cobe_blackbody}
\end{figure}

\begin{figure}[h]
\centering
    \textbf{The Cosmic Microwave Background anisotropies measured by COBE, WMAP and Planck}\par\medskip
\centering
   {\includegraphics[width=.65\textwidth]{cmb1.jpg}}

\caption{(NASA/COBE/DMR; NASA/WMAP SCIENCE TEAM; ESA AND THE PLANCK COLLABORATION)}
\label{cobe_wmap_planck}
\end{figure}

\begin{figure}
\centering
    \textbf{The CMB sky measured by COBE}\par\medskip
\centering
\includegraphics[scale =0.3]{mono_di_cobe}
\caption{The microwave sky as seen by the COBE DMR (differential microwave radiometer) instrument. The top panel shows the microwave sky as seen on a linear tem- perature scale including zero. No anisotropies are visible in this image, because the CMB monopole at 2.725K dominates. In the middle panel, the monopole component has been subtracted. Apart from some slight contamination from the galaxy near the equator (corre- sponding to the plane of our Galaxy), one sees only a nearly perfect dipole pattern, owing to our peculiar motion with respect to the rest frame defined by the CMB. In the bottom panel, both the monopole and dipole components have been removed. Except for the galactic contamination around the equator, one sees the cosmic microwave background anisotropy along with some noise. (Credit: NASA/COBE Science Team)}
\label{cobe_map}
\end{figure}

@@

In this section we explain the meaning of these parameters and what is Universe made up of. 

@@

The CMB appears as an homogeneous map at temperature $T=2.7$ K as shown in the first panel of Figure(\ref{cobe_map}).

@@

In addition we assume that space-time is described by a metric theory of gravity, and we treat space-time as a 4-dimensional manifold characterized by the metric tensor $g_{\mu \nu}$.
The most general metric $g_{\mu \nu}$ satisfying homogeneity and isotropy assumptions is the Friedmann – Lemaître – Robertson – Walker (FLRW) metric written here in terms of the invariant geodesic distance $\dd s ^2 = g_{\mu \nu} \dd x^{\mu} \dd x^{\nu}$:

@@

where $k$ is the space curvature and it determines the geometry of the spatial dimensions, elliptical, Euclidean or hyperbolic for $k>0$, $k=0$ or $k<0$.
The space curvature determines in which kind of three dimensional space we live in and how angles are measured.\\
The scale factor $a(t)$ determines how the distances scale with the evolution of the Universe.
The propagation condition of light is $ds^2 =0$, which implies that the light with frequency $\nu$ and the wavelength $\lambda$ is red- or blueshifted by the same amount as space expands or shrinks:
\dfrac{\nu_{emitted}}{\nu_{observed}} = \dfrac{\lambda_{emitted}}{\lambda_{observed}} = 1 + \dfrac{\lambda_{emitted}-\lambda_{observed}}{\lambda_{observed}}=1+z= \dfrac{a(t_{emitted})}{a(t_{observed})}
\\
Notice that the FLRW metric has $g_{0 i}=0$ due to the isotropy and the two parameters $a(t)$ and $k$ are the only allowed to preserve our assumptions, all the others parameter that we can introduce can be absorbed in these two.

@@

in order to understand how spacetime evolves by using the FLRW metric.
The Einstein's field equations relate how the geometry of spacetime (left hand side) changes with the presence of masses and energy (right hand side).\\

@@

Solving (\ref{einstein_eq})and using the above definition (\ref{energy-momentum-tensor}) (which is a rather long and tedious derivation), gives us the Friedmann equation:

@@

H^2 = \qty(\frac{\dot{a}}{a})^2 &= \dfrac{8 \pi G}{3} \rho -\dfrac{k \, c^2}{a^2} + \frac{\Lambda \, c^2}{3} \\
\vspace{1.5cm}
\dfrac{\ddot{a}}{a} &= -\dfrac{4 \pi G}{3} \qty(\rho + 3 \dfrac{p}{c^2} ) +  \frac{\Lambda \, c^2}{3}

@@

where  $\dot{}$  represents the time derivative.\\
The first Friedmann equation tells us how the Hubble parameter$H(t)$, i.e. the expansion rate,  evolves according to what is contained in our Universe.
The second Friedmann equation tells how the Universe accelerates.\\
In order to solve the system we need another equation: the equation of state $p = w \rho$ that relates the the pressure and the mass density.\\
%
%
%
% comment
%\begin{comment}
By using the conservation of energy momentum tensor $\nabla_\mu T^{\mu 0} =0 $ becomes:

@@

%\end{comment}
%
%
%
% comment
\begin{comment}

@@

 \end{comment}
Solving this for dust (a name for non-relativistic matter) $\rho_m c^2 << p_m$ and for radiation $\rho_r c^2 = p_r/3$ gives

@@

where $\rho_{m 0}$ and $\rho_{r 0}$ are the matter density and radiation density evaluated today $t = t_0$, $a(t_0) = a_0 =1$.
The above result is somehow intuitive because the normal non-relativistic matter is "diluted" as the the spatial distances $a^3$  increase with the expansion of the Universe, i.e. the number of particles per unit volume decreases with the expansion.
The energy density of radiation $u = \rho_r c^2 $ decreases with the power of four because photons experience not only the dilution due to the spatial expansion but their frequencies, accordingly their energies, decreased by the streching of spacetime.\\
We now rewrite the first Friedmann equation as

@@

where we defined $H_0 = H(t_0)$.
The above equation tells us that in the limit of $a \rightarrow 0$ the expansion rate and the evolution of the early Universe was determined by radiation.
In the other limit the future of the Universe is determined by the cosmological constant $\Lambda$.
We have not specified the nature of the cosmological constant, up to now it has been trated as a possible parameter that can be added to the Einstein's field equations(\ref{einstein_eq}).
\\
If we define the density parameters as:

@@

\Omega_{\Lambda 0} = \dfrac{\Lambda c^2}{3 H_0 ^2} \qquad \Omega_{k0} = -  \dfrac{k c^2}{H_0 ^2} \qquad
\Omega_{m0} = \dfrac{8 \pi G}{3 H_0 ^2} \rho_{m 0} \qquad \Omega_{r0} =\dfrac{8 \pi G}{3 H_0 ^2} \rho_{r 0}

@@

 and evaluating the Friedmann equation (\ref{firedmann_eq}) today $a(t_0) =a_0 =1$ we obtain the cosmic sum rule

@@

\Omega_{\Lambda 0} + \Omega_{m 0} + \Omega_{k 0} + \Omega_{r 0} = 1

@@

So the experimental measurements of these parameters must add up to one.
We will see in the next sections how the CMB power spectrum puts constraints on these parameters and tell use that there is a form of matter that we cannot wee.

\section{Phenomenology of the Fluctuations}
The Cosmic Microwave Background fluctuations were produced when radiation and matter were interacting in the early stage of the Universe.
Before the release of the CMB at $z > 1100$ the electromagnetic radiation is firmly coupled (both energetically and kinetically) to the baryons.
The behavior of such a system can be solved by using the equations of relativistic hydrodynamics and the Einstein's field equations (\ref{einstein_eq}) with small perturbations of density $\delta \rho$, pressure $\delta p$ and gravitational potential  $\delta \Phi$.
Perturbations of different scales propagate as sound waves through the plasma until they are frozen at the time of recombination and they appear to us as temperature fluctuations $\Theta = \delta T /T$ in the CMB map.
For a photon gas we know that the temperature fluctuations are related to the number density $n$, energy density $u = \rho c^2$ and pressure $p$ as follow:
\begin{equation}
n \propto T ^3 \rightarrow \dfrac{\delta n}{n_0} = 3 \Theta \qquad \qquad \qquad
 u \propto T^4  \qquad p  = \dfrac{u}{3} \rightarrow \dfrac{\delta u}{u_0} = 4 \Theta = \dfrac{\delta p}{p_0} \notag
\end{equation}
Solving the full general-relativistic description of the photon-matter fluid requires a long and complicated treatment [BAUMANN], so we use instead an analogy with the most favourite model of a physicist: the harmonic oscillator\footnote{As in every analogy in physics, the phenomena are not the same. A mass on a spring is not the same as a relativistic fluid perturbation in an expanding 4-dimensional manifold, but the analogy is very helpful for the understanding and the equations are not that different (in first approximation)}.\\
All kinds of perturbations are proportional to the temperature fluctuations $\Theta$, so we  can associate the fluctuations to the displacement $x$ of spring-mass system. 
The fluctuations of the primordial fluid can be characteried by a typical scale $\lambda =2 \pi /k$ of oscillation/propagation that could depends on the local gravitational pull, pressure or properties. 
In addition, they are damped due to the expasion of the Universe and due to collisions, and ultimately they feel a constant gravitational pull from the overall fluid.
By analogy with the mass-spring system we analyze the fluctuations at different scales by solving the well known equation:
\begin{equation}
m\ddot{x} = - k x - \eta \dot{x} - m g \qquad \rightarrow \qquad \ddot{x} + 2 \gamma \dot{x} + \omega_k ^2 x + g =0
\end{equation}
where $k$ is linked to the scale of oscillations $\omega_k ^2 = k/m$, $\eta$ to the damping ($2 \gamma = \eta /m$), $g$ to the constant gravitational pull of the overall system and $m$ the inertia of the perturbation.
The general solution of the differential equation is 
\begin{equation}
x = x_0 e^{i \, \omega t} - \dfrac{g \, m}{k} \qquad \qquad \omega = i \gamma \pm \sqrt{ \omega_k ^2 - \gamma^2}
\end{equation}
We will discuss how with this simple model we can obtain some of the features of the CMB power spectrum at different scales.

\subsection{Sachs-Wolfe effect}
In the harmonic oscillator model for very small $k$ the mass experiences only the gravitational pull of the overdensities $x \approx - g m/k$ and it does not oscillate.
We can compare the x displacement of the spring with the size of the oscillation $\lambda$
\[
\dfrac{x}{\lambda} \approx -\dfrac{g m}{\lambda k} \dfrac{\lambda}{\lambda} = -\dfrac{g \lambda}{\lambda^2 \omega_k ^2}
\]
Notice that the numerator is the gravitational potential of a uniform field and the denominator is proportional to the speed of the oscillations.\\
In the photon-baryon fluid, perturbation modes of very large scales (large $\lambda$, very small $k$) do not have the time to complete an entire oscillation before recombination and they are influenced only by the gravitational potential of the overdensities.
As a consequence of the Einstein's equivalence principle, photons that escape from a gravitational field are redshifted.
If we consider the fluctuations of the CMB on large scales, the photons that tries to escape from the gravitational well created by matter are redshifted according to the so-called Sachs-Wolfe effect. 
The theoretical result from the relativistic derivation gives a temperature fluctuation $\Theta = -\delta \Phi /(3c^2)$.
The Sachse Wolfe effect takes into account an additional effect that cannot be explained with the simple harmonic oscillator model and it arises from the time dilation due to the expansion of the Universe.

@@

 by starting from the continuity and Euler equation for a photon fluid with  number density $n$, energy density $u$ and pressure $p$:
\begin{equation}
\label{hydro}
\dot{n} + n_0 \vec{\nabla} \cdot \vec{v} = 0 \qquad
\dot{\vec{v}} = - c^2 \dfrac{\vec{\nabla} \delta p}{p_0 + u_0} - \vec{\nabla} \delta \Phi
\end{equation}
where $v$ is the streaming velocity of the perturbations and the subscript $_0$ indicates the background average values of th correspondent quantities.\\

So, the temperature fluctuations are related to the number density, energy density and pressure as follow:
\begin{equation}
\dfrac{\delta n}{n_0} = 3 \Theta \qquad
\dfrac{\delta u}{u_0} = 4 \Theta = \dfrac{\delta p}{p_0}
\end{equation}
We can rewrite eq.(\ref{hydro}) in terms of the temperature fluctuations:
\begin{equation}
\dot{\Theta} + \nabla \cdot \vec{v} = 0 \qquad \dot{\vec{v}} = - c^2 \vec{\nabla} \Theta - \vec{\nabla} \delta \Phi
\end{equation}
where we used $p_0 + u_0 = 4 p_0$

@@

\section{Comments on the results and plots of the CMB spectrum}
We have seen in section 3 how general relativity and the density parameters  characterize the content of our Universe,  in section 4 the dynamics of the CMB photons is affected by matter.
We want to link the previous theoretical results to the observations of the temperature fluctuations of the CMB sky.\\
The temperature fluctuations are projected in the sky map and they appear to us as anisotropies in the sky map tempearature.

The spherical harmonics $Y^{m} _{l} (\theta, \phi)$ compose a set of orthonormal functions on the sky.\\
So if $T(\theta,\phi)$ is the temperature at position $(\theta,\phi)$, it is possible to decompose the temperature into a series:
\begin{equation}
T(\theta, \phi) = \sum _{l, m} a_{lm} Y^{m} _l (\theta, \phi)
\end{equation}
Due to the orthonormality of the spherical harmonics
\begin{equation}
a_{lm} = \int _0 ^{2 \pi} \dd \phi \int _0  ^{\pi} \dd \theta \,
\sin \theta \, T(\theta, \phi) \, Y^{m} _l (\theta, \phi)
 \end{equation}
 We define the power spectrum of the temperature map as 
 \begin{equation}
 C_l = \dfrac{1}{2l+1} \sum _{m=-l} ^{l} |a_{lm}|^2
 \end{equation}
The measured temperature power spectrum is shown in Figure(\ref{temp_pow_spect})
%
%
%
\begin{figure}
\includegraphics[width=0.9\textwidth]{planck2018}
\caption{Planck 2018 temperature power spectrum. $\mathcal{D}^{TT}_l = l(l+1)C_l$ (arXiv:1807.06209v1)}
\label{temp_pow_spect}
\end{figure}
%
%
%
The Sachs-Wolfe effect is dominant on large scales, i.e. small l.\\
The first peak is at $l \approx 200$ and the  acoustic oscillations take place.
The damping effect is evident for $l>10^{3}$.\\
